---
title: "Decimal places for percents in abstracts"
author: "Adrian Barnett"
date: "23 March 2018"
output: word_document
---

```{r setup, include=FALSE}
# to do, plot of decimal places by difference ranges?
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, comment = '', message = FALSE, dpi=300)
options(width=10000)
q1 = function(x){quantile(x, probs=0.25)}
q3 = function(x){quantile(x, probs=0.75)}
library(plyr) # for revalue
library(diagram) # for flow diagram
library(reshape2)
library(doBy) # for summary stats
library(tables) # for nice tables
library(pander) # for nice tables
library(ggplot2) # for nice plots
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", grey(0.4)) # colour-blind palette plus grey

# CIs for multinomial 
source('MultinomialCIsBayes.R')
# already run winbugs? If so, just load saved results
already.run = T
if(already.run==T){load('Multi.results.RData')}

# get the data
load('Analysis.Ready.RData')
# shorter label
levels = unique(data$journal)
labels = c('JAMA','PLOS ONE','BMJ Open','Lancet, HIV','BMJ','EHP','MJA','NEJM','Lancet, P','Lancet, O','F1000','Lancet, D&E','PLOS Med','Lancet, GH','Lancet, RM','Lancet, PH','Lancet','Lancet, H','Lancet, ID','Lancet N','Lancet, G&H','Nature','Lancet, Planet')
data$short.journal <- revalue(data$journal, c(
   "PLoS medicine"                           =  "PLOS Med"          ,
 "PloS one"                                =  "PLOS ONE"      ,
 "Nature"                                  =  "Nature"      ,
 "F1000Research"                           =  "F1000"   ,
 "Lancet (London, England)"                =  "Lancet"           ,
 "The New England journal of medicine"     =  "NEJM"           ,
 "BMJ open"                                =  "BMJ Open"           ,
 "The Lancet. Infectious diseases"         =  "Lancet, ID"          ,
 "The Medical journal of Australia"        =  "MJA"     ,
 "The Lancet. Global health"               =  "Lancet, GH"     ,
 "The Lancet. Oncology"                    =  "Lancet, O"         ,
 "The lancet. Diabetes & endocrinology"    =  "Lancet, D&E"   ,
 "The Lancet. Haematology"                 =  "Lancet, H"      ,
 "BMJ (Clinical research ed.)"             =  "BMJ"    ,
 "Environmental health perspectives"       = "EHP"    ,
 "JAMA"                                    =  "JAMA"    ,
 "The Lancet. Neurology"                   =  "Lancet, N"        ,
 "The lancet. HIV"                         =  "Lancet, HIV"     ,
 "The Lancet. Respiratory medicine"        =  "Lancet, RM"    ,
 "The lancet. Psychiatry"                  =  "Lancet, P"      ,
 "The lancet. Gastroenterology & hepatology"= "Lancet, G&H"   ,
 "The Lancet. Public health"               =  "Lancet, PH"        ,
 "The Lancet. Planetary health"            =  "Lancet, Planet"
 ))
```

## Aim

To evaluate compliance with the guidelines on the number of decimal places to use when presenting percentages in health and medical journals.

## Methods

Percentages were automatically extracted from abstracts and assessed for their compliance with the guidelines.
I chose percentages rather than other statistics, such as the mean, because percentages are easily extractable as they are usually tagged with a '%'.

I chose abstracts as these are freely searchable for all papers, whereas the full text is only searchable for open access publications.
Also, the abstract is a very important summary of the work that will be read more often than the full paper.

The data were accessed on `r format(date.extracted, '%Y-%b-%d')`.

I ignored any negative signs as the focus was on the decimal places rather than the actual value. I also ignored any mathematical symbols, so for example, "<1%" would be recorded as 1 percent with no decimal places. For example, "incident tuberculosis occurred in six (<1%) of 795 contacts" 29273539[PMID] (PMID = pubmed ID).

## Results

### Flow chart of abstracts selected

```{r}
n.search = 23293 + 9854 # PLOS + original search
n.articles = 22651 + 9243 
n.oneperc = length(unique(data$pubmed))
n.percents = nrow(data)
#
flow = function(){ # use as a function so that it can be repeated for jpeg
  l1 = paste('Papers published in selected\njournals in 2017 (n=',format(n.search, big.mark=','),')', sep='')
  l2 = paste('Papers that were research\narticles (n=',format(n.articles, big.mark=','),')', sep='')
  l3 = paste('Papers with at least one\npercent in abstract (n=',format(n.oneperc, big.mark=','),')', sep='')
  l4 = paste('Total number of\npercents (n=',format(n.percents, big.mark=','),')', sep='')
  labels = c(l1,l2,l3,l4)
  n.labs = length(labels)
  M <- matrix(nrow = n.labs, ncol = n.labs, byrow = TRUE, data = 0)
  M[2,1] = M[3,2] = M[4,3] =  "' '"
  pos = matrix(data=c(0.5,0.9,
                      0.5,0.63,
                      0.5,0.36,
                      0.5,0.1), ncol=2, byrow = T)
  par(mai=rep(0.01,4))
  plotmat(M, pos=pos, name=labels, main=NULL, shadow.size = 0, curve = 0, box.type = 'square', 
          cex.txt = 0.7, box.size = 0.29, box.prop = 0.2)  
}
flow()
# export to figure for paper
jpeg('figure1.jpg', units='in', res=400, width=4, height=5, quality=100)
flow()
invisible(dev.off())
```


### Summary statistics on the number of percents per abstract

```{r summary.stats}
tab = data.frame(table(data$pubmed))
tab = tabular(Heading('Number of percents')*Freq ~ (n=1) + mean + median + q1 + q3 + min + max, data=tab)
pander(tab, style='simple', digits = 2, emphasize.rownames = F)
```

The median number of percents per abstract was 4, with an inter-quartile range of 2 to 8. These statistics exclude abstracts with no percentages.



### Histogram of all percents in integer bins

```{r outside, include=F}
outside = sum(data$percents>100)
outside.percent = round(1000*(outside/nrow(data)))/100
```

The histogram excludes the `r outside` (`r outside.percent`%) percents greater than 100%.

```{r histogram}
to.plot = subset(data, percents < 100.5)
hplot = ggplot(data=to.plot, aes(x=percents))+
  geom_histogram(breaks=seq(-0.5, 100.5, 1))+
  theme_bw()+
  theme(panel.grid.minor = element_blank())+
  xlab('Observed percent')+
  ylab('Frequency')+
  scale_x_continuous(breaks=seq(0,101,10))
hplot
# export to figure for paper
jpeg('figure2.jpg', units='in', res=400, width=5, height=4, quality = 100)
print(hplot)
invisible(dev.off())
```

There was clearly digit preference at multiples of 10%, particularly 50%, and at 75% and 95%. This was because there was some use of percentages as reference points, for example, 

* "If screening coverage fell to 50%" in 29307386[PMID].
* "and more than 50% of the suicides occurred in females" 28716352[PMID] 

Same plot coloured by number of decimal places, or facetted by number of decimal places.

### Number of abstracts where all percents follow the guidelines

```{r all.correct}
data$correct = as.numeric(data$diff==0)
all.correct = summaryBy(correct ~ pubmed, FUN=max, data=data)
all.correct$correct.max = factor(all.correct$correct.max, levels=0:1, labels=c('Yes','No'))
tab = tabular(Heading('All percents correct')*correct.max + 1 ~ (n=1) + Percent("col"), data=all.correct)
pander(tab, style='simple', digits = 2, emphasize.rownames = F)
```

The table shows the number of abstracts where all percents were correct.

### Numbers of decimal places

```{r decimals}
tab = tabular(Heading('Number of decimal places')*factor(observed.dps) + 1~ (n=1) + Percent("col"), data=data)
pander(tab, style='simple', digits = c(0,0,0), emphasize.rownames = F)
```

Most decimal places were less than 2.

### Numbers of significant figures

```{r sf}
tab = tabular(Heading('Number of significant figures')*factor(observed.sfs) + 1~ (n=1) + Percent("col"), data=data)
pander(tab, style='simple', digits = c(0,0,0), emphasize.rownames = F)
```

### Table of differences between observed and ideal decimal places

```{r diff.table}
tab = tabular(Heading('Difference')*factor(diff) + 1~ (n=1) + Percent("col"), data=data)
pander(tab, style='simple', digits = c(0,0,0), emphasize.rownames = F)
```

Negative differences are too few digits, positive difference are too many.


### Percents correct by journal (plot)

The rows in the plot are ordered by the percent ``Just right''.

```{r journal.percent, fig.width=6, fig.height=6}
data$diff = sign(data$diff) # tranform to negative, zero or positive
# numbers by journal 
tab.nums = with(data, table(short.journal, diff))
# percents by journal 
tab = prop.table(tab.nums, 1) *100
to.plot.percent = data.frame(tab)
to.order = subset(to.plot.percent, diff== 0) # by most correct
to.order = to.order[order(to.order$Freq),]
levels =  to.order$short.journal
to.plot.percent$short.journal = factor(as.character(to.plot.percent$short.journal), levels=levels)
# use factor for ordering
to.plot.percent$diff = factor(to.plot.percent$diff, levels=c(1,0,-1), ordered=T)
to.plot.percent = to.plot.percent[order(-as.numeric(to.plot.percent$diff)),]
bplot = ggplot(data=to.plot.percent, aes(x=factor(short.journal), y=Freq, fill=factor(diff)))+
  geom_bar(stat= "identity")+
  theme_bw()+
  theme(legend.position = 'top', panel.grid.minor = element_blank(), plot.margin=unit(c(1,1,1,1),"mm"))+
  xlab('')+
  ylab('Percent')+
  scale_fill_manual('Decimal\nplaces', breaks=c(-1,0,1), labels=c('Too few','Just right','Too many'), values=cbbPalette[2:4])+
  coord_flip()
bplot
```

### Percents correct by journal (table)

Formatted for reading into Word.

```{r journal.table, results='asis'}
to.table = to.plot.percent # percents
long = melt(tab.nums) # numbers
for.table = merge(to.table, long, by=c('short.journal','diff')) # merge
index = for.table$Freq<10 | for.table$Freq>90
for.table$Freq[index] = round(for.table$Freq[index],1) # great irony here
for.table$Freq[!index] = round(for.table$Freq[!index],0) 
journal.list = to.order$short.journal # use this ordering
cat('Journal&Too few&Just right&Too many&Total\n', sep='') # header
for (j in journal.list){
  this.few = subset(for.table, short.journal==j & diff == -1)
  this.ideal = subset(for.table, short.journal==j & diff ==  0)
  this.many = subset(for.table, short.journal==j & diff ==  1)
  Total = this.few$value + this.ideal$value + this.many$value
  cat(j, ':', this.few$Freq, '% (', this.few$value, ')', 
         ':', this.ideal$Freq, '% (', this.ideal$value, ')',
         ':', this.many$Freq, '% (', this.many$value, ')', 
         ':', Total, '\n', sep='')
}
```

### Confidence intervals for overall numbers

Assuming a Bayesian Multinomial Dirichlet model.

```{r ci, results='asis'}
tab = table(data$diff)
if(already.run==F){
  this = multiCIs(tab)
  multi.res = list()
  multi.res[[1]] = this
}
if(already.run==T){
  load('Multi.results.RData')
}
pander(multi.res[[1]], style='simple', emphasize.rownames = F)
```

### Histogram of observed range in percents from the same sentence or abstract (base 2 log-transformed)

```{r diff}
to.plot = subset(data, is.infinite(range.diff)==F)
hplot = ggplot(data=to.plot, aes(x=log2(range.diff+1)))+
  geom_histogram()+
  theme_bw()+
  ylab('Frequency')
hplot
```



### Plot types of incorrect

The plot is a summary for each journal. The plot includes 95\% confidence intervals for both percents to show the uncertainty for each journal.

```{r crossplot, fig.width=6, fig.height=5}
# numbers
tab = with(data, table(short.journal, diff))
to.plot.nums = data.frame(tab)
# 
percent.wide = dcast(subset(to.plot.percent), short.journal ~ diff)
names(percent.wide) = c('journal','many','okay','few')
nums.wide = dcast(subset(to.plot.nums), short.journal ~ diff)
names(nums.wide) = c('journal','many','okay','few')
#nums.wide = subset(nums.wide, journal != 'The Lancet. Planetary health') # too few
percent.wide$xlower = percent.wide$xupper = percent.wide$ylower = percent.wide$yupper =NA
if(already.run==F){
  for (k in 1:nrow(percent.wide)){
    this.journal = percent.wide$journal[k]
    mnums = as.numeric(nums.wide[which(nums.wide$journal==this.journal), 2:4])
    m = multiCIs(mnums)
    percent.wide$xlower[k] = m$lower[1]*100
    percent.wide$xupper[k] = m$upper[1]*100
    percent.wide$ylower[k] = m$lower[3]*100
    percent.wide$yupper[k] = m$upper[3]*100
  }
  multi.res[[2]] = percent.wide
}
if(already.run==T){
  load('Multi.results.RData')
  percent.wide = multi.res[[2]]
}

# plot
percent.wide$cols = 1
percent.wide$cols[grep('lancet', tolower(percent.wide$journal))] = 2
percent.wide$cols[grep('bmj', tolower(percent.wide$journal))] = 3
percent.wide$cols[grep('plos', tolower(percent.wide$journal))] = 4
gplot = ggplot(data=percent.wide, aes(x=few, y=many, xmin=xlower, xmax=xupper, ymin=ylower, ymax=yupper, col=factor(cols)))+
  geom_point()+
  geom_errorbarh()+
  geom_errorbar()+
  scale_color_manual('Journal group', values=cbbPalette[2:5], labels=c('Other','Lancet','BMJ','PLOS'))+
  theme_bw()+
  scale_x_continuous(limits=c(0,NA))+
  scale_y_continuous(limits=c(0,NA))+
  xlab('Too few digits (%)')+
  ylab('Too many digits (%)')+
  geom_point(aes(x=0,y=0), col='black', pch=19, size=2)+
  geom_text(aes(x=0,y=2), label='Ideal', col='black')+
  theme(legend.position = c(0.8,0.8), panel.grid.minor = element_blank())
#  geom_label(aes(x=few, y=many, label=short.journal))
gplot
# export to figure for paper
jpeg('figure3.jpg', units='in', res=500, width=5, height=4, quality = 100)
print(gplot)
invisible(dev.off())
```

#### Alternative plot including journal names

```{r alt.plot, fig.width=6, fig.height=5}
gplot2 = gplot + geom_text(aes(x=few, y=many, label=journal))
gplot2
```

### Sensitivity analysis excluding percents from 90 to 100 and digit preference percents, 10, 20, 30, 40, 50, 60, 70, 75, 80

```{r table.sensitivity}
## exclude 90 to 100
index = data$percents>=90 & data$percents<=100
sub = data[!index, ]
## exclude 90 to 100
to.exclude = c(seq(10,80,10),75)
index = sub$percents %in% to.exclude & sub$observed.dps==0
sub = sub[!index, ]
## recreate numbers
tab = with(sub, table(diff))
if(already.run==F){
  this = multiCIs(tab)
  multi.res[[3]] = this
}
if(already.run==T){
  load('Multi.results.RData')
}
pander(multi.res[[3]], style='simple', emphasize.rownames = F)
```

The table shows the new percents for all journals.

```{r crossplot.sensitivity, fig.width=6, fig.height=5}
tab = with(sub, table(diff, short.journal)) # table by journal
to.plot.nums = data.frame(tab)
# percents by journal (use previous table)
tab = prop.table(tab, 2) *100
to.plot.percent = data.frame(tab)
# transform to wide
percent.wide.sens = dcast(subset(to.plot.percent), short.journal ~ diff)
names(percent.wide.sens) = c('journal','few','okay','many')
nums.wide = dcast(subset(to.plot.nums), short.journal ~ diff)
names(nums.wide) = c('journal','few','okay','many')
percent.wide.sens$xlower = percent.wide.sens$xupper = percent.wide.sens$ylower = percent.wide.sens$yupper =NA
if(already.run==F){
  for (k in 1:nrow(percent.wide.sens)){
    this.journal = percent.wide.sens$journal[k]
    mnums = as.numeric(nums.wide[which(nums.wide$journal==this.journal), 2:4])
    m = multiCIs(mnums);
    percent.wide.sens$xlower[k] = m$lower[1]*100
    percent.wide.sens$xupper[k] = m$upper[1]*100
    percent.wide.sens$ylower[k] = m$lower[3]*100
    percent.wide.sens$yupper[k] = m$upper[3]*100
  }
  multi.res[[4]] = percent.wide.sens
  # save multinomial results
  save(multi.res, file='Multi.results.RData')
}
if(already.run==T){
  load('Multi.results.RData')
  percent.wide.sens = multi.res[[4]]
}

# plot
percent.wide.sens$cols = 1
percent.wide.sens$cols[grep('lancet', tolower(percent.wide.sens$journal))] = 2
percent.wide.sens$cols[grep('bmj', tolower(percent.wide.sens$journal))] = 3
percent.wide.sens$cols[grep('plos', tolower(percent.wide.sens$journal))] = 4
gplot = ggplot(data=percent.wide.sens, aes(x=few, y=many, xmin=xlower, xmax=xupper, ymin=ylower, ymax=yupper, col=factor(cols)))+
  geom_point()+
  geom_errorbarh()+
  geom_errorbar()+
  scale_color_manual('Journal group', values=cbbPalette[2:5], labels=c('Other','Lancet','BMJ','PLOS'))+
  theme_bw()+
  scale_x_continuous(limits=c(0,NA))+
  scale_y_continuous(limits=c(0,NA))+
  xlab('Too few digits (%)')+
  ylab('Too many digits (%)')+
  geom_point(aes(x=0, y=0), col='black', pch=19, size=2)+
  geom_text(aes(x=0, y=2), label='Ideal', col='black')+
  theme(legend.position = c(0.8,0.8), panel.grid.minor = element_blank())
gplot
```

### Movement between two previous plots

The arrows show how journals moved between the full analysis and the sensitivity analysis excluding the digit preference results.

```{r movement, fig.width=6, fig.height=5}
one = subset(percent.wide, select=c('journal','few','many'))
names(one) = c('journal','x0','y0')
two = subset(percent.wide.sens, select=c('journal','few','many','cols'))
names(two) = c('journal','x1','y1','cols')
three = merge(one, two, by='journal')
par(mai=c(0.9,0.9,0.1,0.1), las=1)
maxx = max(c(three$x0, three$x1))
maxy = max(c(three$y0, three$y1))
#plot(0:50, 0:50, xlab='Too few digits(%)', ylab='Too many digits(%)', type='n')
plot(three$x0, three$y0, xlab='Too few digits(%)', ylab='Too many digits(%)', type='p', col=cbbPalette[three$cols+1], xlim = c(0,maxx), ylim=c(0,maxy))
points(x=0, y=0, pch=19)
text(x=0.5, y=2, 'Ideal')
for(k in 1:nrow(three)){
  arrows(x0=three$x0[k], x1=three$x1[k], y0=three$y0[k], y1=three$y1[k], length=0.1, col=cbbPalette[three$cols[k]+1], lwd=1.2)
}
```

### Look for threshold in each journal

```{r table}
data$below = factor(data$percents < 10)
tab = tabular(Heading('Below 10%')*below ~ factor(diff)*((n=1)+Percent('row')), data=data)
pander(tab, digits = 2, emphasize.rownames = F)
```

```{r threshold}
library(splines)
sub = subset(data, journal=='Nature' & percents<100.5)
model = glm(I(diff==0) ~ ns(percents,3), data, family=binomial())
plot = ggplot(data=sub, aes(x=factor(diff), y=percents))+
  geom_boxplot()+
  theme_bw()
plot
```


Problem: 29276017[pmid] has 95% with no explanation (i.e., no ``CI or PI''); hard to program.



